XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-24 talk/icooolps2011/paper.tex
- improve fixed-width font - turn some of the code into figures

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage[utf8]{inputenc}

\makeatletter
\def\PY@reset{\let\PY@it=\relax \let\PY@bf=\relax%
    \let\PY@ul=\relax \let\PY@tc=\relax%
    \let\PY@bc=\relax \let\PY@ff=\relax}
\def\PY@tok#1{\csname PY@tok@#1\endcsname}
\def\PY@toks#1+{\ifx\relax#1\empty\else%
    \PY@tok{#1}\expandafter\PY@toks\fi}
\def\PY@do#1{\PY@bc{\PY@tc{\PY@ul{%
    \PY@it{\PY@bf{\PY@ff{#1}}}}}}}
\def\PY#1#2{\PY@reset\PY@toks#1+\relax+\PY@do{#2}}

\def\PY@tok@gd{\def\PY@bc##1{\fcolorbox[rgb]{0.80,0.00,0.00}{1.00,0.80,0.80}{##1}}}
\def\PY@tok@gu{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.00}{##1}}}
\def\PY@tok@gt{\def\PY@tc##1{\textcolor[rgb]{0.60,0.80,0.40}{##1}}}
\def\PY@tok@gs{\let\PY@bf=\textbf}
\def\PY@tok@gr{\def\PY@tc##1{\textcolor[rgb]{1.00,0.00,0.00}{##1}}}
\def\PY@tok@cm{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
\def\PY@tok@vg{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
\def\PY@tok@m{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@mh{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@cs{\let\PY@bf=\textbf\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
\def\PY@tok@ge{\let\PY@it=\textit}
\def\PY@tok@vc{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
\def\PY@tok@il{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@go{\def\PY@tc##1{\textcolor[rgb]{0.67,0.67,0.67}{##1}}}
\def\PY@tok@cp{\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,0.60}{##1}}}
\def\PY@tok@gi{\def\PY@bc##1{\fcolorbox[rgb]{0.00,0.80,0.00}{0.80,1.00,0.80}{##1}}}
\def\PY@tok@gh{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.00}{##1}}}
\def\PY@tok@ni{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,0.60}{##1}}}
\def\PY@tok@nl{\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,1.00}{##1}}}
\def\PY@tok@nn{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.80,1.00}{##1}}}
\def\PY@tok@no{\def\PY@tc##1{\textcolor[rgb]{0.20,0.40,0.00}{##1}}}
\def\PY@tok@na{\def\PY@tc##1{\textcolor[rgb]{0.20,0.00,0.60}{##1}}}
\def\PY@tok@nb{\def\PY@tc##1{\textcolor[rgb]{0.20,0.40,0.40}{##1}}}
\def\PY@tok@nc{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.67,0.53}{##1}}}
\def\PY@tok@nd{\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,1.00}{##1}}}
\def\PY@tok@ne{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.80,0.00,0.00}{##1}}}
\def\PY@tok@nf{\def\PY@tc##1{\textcolor[rgb]{0.80,0.00,1.00}{##1}}}
\def\PY@tok@si{\def\PY@tc##1{\textcolor[rgb]{0.67,0.00,0.00}{##1}}}
\def\PY@tok@s2{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@vi{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
\def\PY@tok@nt{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.20,0.00,0.60}{##1}}}
\def\PY@tok@nv{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
\def\PY@tok@s1{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@gp{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,0.60}{##1}}}
\def\PY@tok@sh{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@ow{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,0.00}{##1}}}
\def\PY@tok@sx{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@bp{\def\PY@tc##1{\textcolor[rgb]{0.20,0.40,0.40}{##1}}}
\def\PY@tok@c1{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
\def\PY@tok@kc{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@c{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
\def\PY@tok@mf{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@err{\def\PY@tc##1{\textcolor[rgb]{0.67,0.00,0.00}{##1}}\def\PY@bc##1{\colorbox[rgb]{1.00,0.67,0.67}{##1}}}
\def\PY@tok@kd{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@ss{\def\PY@tc##1{\textcolor[rgb]{1.00,0.80,0.20}{##1}}}
\def\PY@tok@sr{\def\PY@tc##1{\textcolor[rgb]{0.20,0.67,0.67}{##1}}}
\def\PY@tok@mo{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@mi{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@kn{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@o{\def\PY@tc##1{\textcolor[rgb]{0.33,0.33,0.33}{##1}}}
\def\PY@tok@kr{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@s{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@kp{\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@w{\def\PY@tc##1{\textcolor[rgb]{0.73,0.73,0.73}{##1}}}
\def\PY@tok@kt{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.47,0.53}{##1}}}
\def\PY@tok@sc{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@sb{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@k{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@se{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@sd{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}

\def\PYZbs{\char`\\}
\def\PYZus{\char`\_}
\def\PYZob{\char`\{}
\def\PYZcb{\char`\}}
\def\PYZca{\char`\^}
% for compatibility with earlier versions
\def\PYZat{@}
\def\PYZlb{[}
\def\PYZrb{]}
\makeatother

@@

\section{Introduction}

\section{The PyPy Project}
\label{sect:pypy}

@@

\section{Tracing JIT Compilers}
\label{sect:tracing}

\section{Controlling The Extent of Tracing}

The question I was asked most often during my recent \href{http://morepypy.blogspot.com/2011/03/us-trip-report-popl-microsoft-ibm.html}{US trip} was how exactly
the hints work that interpreter authors can use to improve the execution speed
of the programs running on their interpreters. Since those hints are not really
documented all that well, I decided to write blog posts about them. This is the
first one.

@@

As was explained in a \href{http://morepypy.blogspot.com/2009/03/applying-tracing-jit-to-interpreter.html}{blog post} and a \href{http://codespeak.net/svn/pypy/extradoc/talk/icooolps2009/bolz-tracing-jit.pdf}{paper} two years ago, PyPy's JIT is a
meta-tracer. Since we want to re-use our tracer for a variety of languages, we
don't trace the execution of the user program, but instead trace the execution
of the \emph{interpreter} that is running the program. This means that the traces
don't contain the bytecodes of the language in question, but RPython-level
operations that the interpreter did to execute the program.

@@

If the interpreter author finds false negatives or false positives, she can fix
that by applying a hint to the tracer. These hints take the form of function
decorators (which both live in the \texttt{pypy.rlib.jit} module). In the next two
subsections I will describe these two function decorators and their use.

@@

@@

The second decorator \texttt{dont\_look\_inside} is used to fix false positives. It
tells the JIT to never trace into the decorated function and just always produce
a residual call instead. This decorator is in many ways less important than the
unrolling one (except for a special situation that I will describe in a
follow-up post). It is used if tracing into a function is not expected to yield
any speed benefits, because the optimizer will not be able to improve it much.
This is often the case if the called helper function does not contain any
``dynamic'' behaviour. In such a situation it is better to just leave the function
call in the trace, because that produces less code.

In this post we discussed two hints that can be used to control precisely which
parts of the interpreter should be meta-traced. If these hints are used
carefully, this can go a long way to making the interpreter produce traces that
contain exactly the interesting part of the execution, and will contain calls to
the functions that can not be optimized by tracing techniques.

In the next part of this series I will discuss a different set of hints that can
be used to strongly optimize traces.

% Document title
\section{Controlling the Tracing of an Interpreter With Hints, Part 2: Controlling Optimization}

This is part 2 of a series on how to speed up an interpreter written with PyPy
by adding JIT hints to the interpreter. Part 1 described how to \href{http://morepypy.blogspot.com/2011/03/controlling-tracing-of-interpreter-with.html}{control the
extent of tracing}. In this post I will describe how to add hints that
influence the optimizer.  If applied correctly these techniques can give
really big speedups by pre-computing parts of what happens at runtime. On the other
hand, if applied incorrectly they might lead to code bloat, thus making the
resulting program actually slower.

@@

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or reduce
the number of operations: most of these are well known \href{http://en.wikipedia.org/wiki/Compiler_optimization\#Optimization_techniques}{compiler optimization
techniques}, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques:
%
\begin{itemize}

\item \href{http://en.wikipedia.org/wiki/Constant_folding}{constant folding}

\item \href{http://en.wikipedia.org/wiki/Common_subexpression_elimination}{common subexpression elimination}

\item allocation removal, as described in the paper that I recently \href{http://morepypy.blogspot.com/2011/03/us-trip-report-popl-microsoft-ibm.html}{presented at
PEPM}

\item store/load propagation

\item \href{http://morepypy.blogspot.com/2011/01/loop-invariant-code-motion.html}{loop invariant code motion}

@@

In this post I will describe two hints that allow the interpreter author to
increase the optimization opportunities for constant folding. For constant
folding to work, two conditions need
to be met:
%
\begin{itemize}

\item the arguments of an operation actually need to all be constant,
i.e. statically known by the optimizer

\item the operation needs to be \emph{pure}, i.e. always yield the same result given
the same arguments.

@@

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function. Even
though in principle the argument to a Python function could be any Python type,
in practice the argument types tend to not vary often. Therefore it is possible to
promote the types. In the next blog post I will give a complete example of how
this works.

@@

@@

In this blog post I explained two more hints that can be used in the source code
of the interpreter. They are used to influence what the optimizer does with the
trace. I realize the examples given here are a bit too small, in the next
installment I will give a worked-out example that puts all the pieces together.

\section{Controlling the Tracing of an Interpreter With Hints, Part 3: Putting Things Together}

This is part 3 of the series on how to speed up an interpreter written with
PyPy by adding JIT hints to the interpreter. Part 1 described how to \href{http://morepypy.blogspot.com/2011/03/controlling-tracing-of-interpreter-with.html}{control
the extent of tracing}. Part 2 described how to \href{http://morepypy.blogspot.com/2011/03/controlling-tracing-of-interpreter-with_15.html}{influence the optimizer with
promotion and pure functions}. In this post I describe a worked-out example of
a small object model for a dynamic language and how to make it efficient using
the hints described in the previous posts.

@@

For the purpose of this blog post we will use a very simple and bare-bones
object model that just supports very simple classes and instances, without any
inheritance or any fancy features. The model has classes, which contain methods.
Instances have a class. Instances have their own attributes. When looking up an
attribute on an instance, the instances attributes are searched. If the
attribute is not found there, the class' attributes are searched.

To implement this object model, we could use the following RPython code as part
of the interpreter source code:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{name} \PY{o}{=} \PY{n}{name}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{k}{def} \PY{n+nf}{instantiate}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n}{Instance}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
            \PY{k}{return} \PY{n}{result}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}

\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
            \PY{k}{return} \PY{n}{result}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}

    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{try}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
\end{Verbatim}

@@

The trace could look like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{attributes1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{attributes}
\PY{n}{result1} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{attributes1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result1} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

\PY{c}{# inst.getattr("b")}
\PY{n}{attributes2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{attributes}
\PY{n}{v1} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{attributes2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{v1} \PY{o+ow}{is} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods1} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
\PY{n}{result2} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}

\PY{c}{# inst.getattr("c")}
\PY{n}{attributes3} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{attributes}
\PY{n}{v3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{attributes3}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{v3} \PY{o+ow}{is} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods2} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
\PY{n}{result3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

@@

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints we have looked at in the
two earlier blog posts don't seem to help with the current object model. There is
no pure function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

@@

Therefore it makes sense to factor the layout information out of the instance
implementation into a shared object. This shared layout object is called a
\emph{map}. Maps are an old idea that comes originally from the SELF language. They are
also used by many JavaScript implementations such as V8. I've \href{http://morepypy.blogspot.com/2010/11/efficiently-implementing-python-objects.html}{written about maps
before}, so I won't explain them fully again.

The rewritten \texttt{Instance} class using maps looks like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{new\PYZus{}map\PYZus{}with\PYZus{}additional\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{if} \PY{n}{name} \PY{o+ow}{not} \PY{o+ow}{in} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{:}
            \PY{n}{newmap} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
            \PY{n}{newmap}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{o}{.}\PY{n}{update}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{p}{)}
            \PY{n}{newmap}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n+nb}{len}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{p}{)}
            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{newmap}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]}

\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}

\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}

    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage[scaled=0.8]{beramono}
\usepackage[utf8]{inputenc}

\input{code/style.tex}

@@

%___________________________________________________________________________
\section{Introduction}

%___________________________________________________________________________
\section{The PyPy Project}
\label{sect:pypy}

XXX
\cite{armin_rigo_pypys_2006}

%___________________________________________________________________________
\section{Tracing JIT Compilers}
\label{sect:tracing}

@@

%___________________________________________________________________________
\section{Controlling The Extent of Tracing}

XXX how exactly
the hints work that interpreter authors can use to improve the execution speed
of the programs running on their interpreters?

@@

XXX As was explained in a \href{http://morepypy.blogspot.com/2009/03/applying-tracing-jit-to-interpreter.html}{blog post} and a \href{http://codespeak.net/svn/pypy/extradoc/talk/icooolps2009/bolz-tracing-jit.pdf}{paper} two years ago, PyPy's JIT is a
meta-tracer. Since we want to re-use our tracer for a variety of languages, we
don't trace the execution of the user program, but instead trace the execution
of the \emph{interpreter} that is running the program. This means that the traces
don't contain the bytecodes of the language in question, but RPython-level
operations that the interpreter did to execute the program.

@@

If the interpreter author finds false negatives or false positives, she can fix
that by applying a hint to the tracer. These hints take the form of function
decorators (which both live in the \Verb|pypy.rlib.jit| module). In the next two
subsections we describe these two function decorators and their use.

@@

The second decorator \texttt{dont\_look\_inside} is used to fix false positives. It
tells the JIT to never trace into the decorated function and just always produce
a residual call instead. This decorator is in many ways less important than the
unrolling one (except for a special situation that is described in
Section XXX). It is used if tracing into a function is not expected to yield
any speed benefits, because the optimizer will not be able to improve it much.
This is often the case if the called helper function does not contain any
``dynamic'' behaviour. In such a situation it is better to just leave the function
call in the trace, because that produces less code.

In this section we discussed two hints that can be used to control precisely which
parts of the interpreter should be meta-traced. If these hints are used
carefully, this can go a long way to making the interpreter produce traces that
contain exactly the interesting part of the execution, and will contain calls to
the functions that can not be optimized by tracing techniques.

In the next section we discuss a different set of hints that can
be used to strongly optimize traces.

@@

\section{Controlling Optimization}

The last section described how to control the
extent of tracing. In this section we will describe how to add hints that
influence the optimizer.  If applied correctly these techniques can give
really big speedups by pre-computing parts of what happens at runtime. On the other
hand, if applied incorrectly they might lead to code bloat, thus making the
resulting program actually slower.

@@

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or reduce
the number of operations: most of these are well known \href{http://en.wikipedia.org/wiki/Compiler_optimization\#Optimization_techniques}{compiler optimization
techniques}, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques:
%
\begin{itemize}
    \item constant folding
    \item common subexpression elimination
    \item allocation removal \cite{bolz_allocation_2011}
    \item store/load propagation
    \item loop invariant code motion
\end{itemize}

@@

In this section we describe two hints that allow the interpreter author to
increase the optimization opportunities for constant folding. For constant
folding to work, two conditions need
to be met:
%
\begin{itemize}
    \item the arguments of an operation actually need to all be constant,
    i.e. statically known by the optimizer
    \item the operation needs to be \emph{pure}, i.e. always yield the same result given
    the same arguments.
\end{itemize}

@@

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function. Even
though in principle the argument to a Python function could be any Python type,
in practice the argument types tend to not vary often. Therefore it is possible to
promote the types. The next section will present a complete example of how
this works.

@@

In this section we presented two more hints that can be used in the source code
of the interpreter. They are used to influence what the optimizer does with the
trace. The examples given here are a bit too small, the next
section gives a worked-out example that puts all the pieces together.

@@

\section{Putting Things Together}

In this section we describe a worked-out example of
a small object model for a dynamic language and how to make it efficient using
the hints described in the previous sections.

@@

For the purpose of this section we will use a very simple and bare-bones
object model that just supports very simple classes and instances, without any
inheritance or any fancy features. The model has classes, which contain methods.
Instances have a class. Instances have their own attributes. When looking up an
attribute on an instance, the instances attributes are searched. If the
attribute is not found there, the class' attributes are searched.

To implement this object model, we could use the following RPython code as part
of the interpreter source code:

\begin{figure}
\input{code/interpreter-slow.tex}
\caption{Original Version of a Simple Object Model}
\label{fig:interpreter-slow}
\end{figure}

@@

The trace could look like this:

\begin{figure}
\input{code/trace1.tex}
\caption{Trace Through the Object Model}
\label{fig:trace1}
\end{figure}

@@

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints we have looked at in the
two previous sections don't seem to help with the current object model. There is
no pure function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

@@

Therefore it makes sense to factor the layout information out of the instance
implementation into a shared object. This shared layout object is called a
\emph{map}. Maps are an old idea that comes originally from the SELF language \cite{XXX}. They are
also used by many JavaScript implementations such as V8.

The rewritten \texttt{Instance} class using maps looks like this:

\begin{figure}
\input{code/map.tex}
\caption{Simple Object Model With Maps}
\label{fig:maps}
\end{figure}
